---
title: "Football Performance Monitoring — R Analysis"
subtitle: "Master 2SIA — Python-R-Git Project"
author: "Celine BENDEKOUM"
date: "02/04/2026"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    code_folding: show
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 9,
  fig.height = 5,
  fig.align  = "center"
)
```

# Required libraries

```{r libraries}
# Data manipulation
library(readr)
library(dplyr)
library(tidyr)

# Visualization
library(ggplot2)
library(scales)

# Statistical tests
library(coin)       # non-parametric tests
library(rstatix)    # tidy statistical summaries

# Table display
library(knitr)
library(kableExtra)
```

---

# Scientific question

A football club wants to **monitor the physical load** of its players
across matches and training sessions using GPS/IMU data.

The R analysis focuses on three complementary questions:

1. **Is the physical load significantly different between matches and
   training sessions?** (per KPI, per position)
2. **How does load evolve over time?** (trends, peaks, overload
   detection)
3. **Can a simple linear model predict distance from speed/acceleration
   variables?** (and are the assumptions verified?)

---

# Data

## Loading

The raw file is `data/data/summary.csv`. All paths are **relative** so
the project works on any computer.

```{r load-data}
raw <- read_csv(
  "data/data/summary.csv",
  na      = c("NA", "NaN", "N/A", ""),
  show_col_types = FALSE
)

# Standardise column names and select variables of interest
df_raw <- raw |>
  rename(
    sensor        = Sensor,
    last_name     = last_name,
    first_name    = first_name,
    session_ts    = date,
    session_type  = type,
    position      = position,
    distance_km   = feature_distance,
    avg_speed     = feature_avg_speed,
    max_speed     = feature_max_speed,
    max_acc       = feature_max_acc,
    d_0_5         = feature_d_speed_0_5,
    d_5_10        = feature_d_speed_5_10,
    d_10_15       = feature_d_speed_10_15,
    d_15_21       = feature_d_speed_15_21,
    d_21_24       = feature_d_speed_21_24,
    d_24_30       = feature_d_speed_24_30,
    d_over30      = feature_d_speed_over30,
    acc_3mss      = feature_acc_3mss,
    dec_3mss      = feature_dec_3mss
  ) |>
  mutate(
    session_ts = as.POSIXct(session_ts),
    session_date = as.Date(session_ts)
  )

glimpse(df_raw)
```

## Quality control & cleaning

We apply the same rules as the Python pipeline so results are
comparable.

```{r cleaning}
# Constants (same as Python)
MIN_MATCH_DIST <- 0.5
MAX_MATCH_DIST <- 20.0
MAX_MAX_SPEED  <- 40.0
ZONE_TOL       <- 1.15

df <- df_raw |>
  filter(
    session_type %in% c("game", "practice"),
    !is.na(session_ts),
    !is.na(sensor)
  ) |>
  mutate(
    zones_sum = rowSums(
      across(c(d_0_5, d_5_10, d_10_15, d_15_21, d_21_24, d_24_30, d_over30)),
      na.rm = TRUE
    )
  ) |>
  filter(
    # No negative values
    is.na(distance_km) | distance_km >= 0,
    is.na(max_speed)   | max_speed   >= 0,
    is.na(acc_3mss)    | acc_3mss    >= 0,
    is.na(dec_3mss)    | dec_3mss    >= 0,
    # Zone coherence
    is.na(distance_km) | zones_sum <= distance_km * ZONE_TOL,
    # Match-specific rules
    session_type != "game" | (
      distance_km >= MIN_MATCH_DIST &
      distance_km <= MAX_MATCH_DIST &
      (is.na(max_speed) | max_speed <= MAX_MAX_SPEED)
    )
  )

# Summary of what remains
df |>
  count(session_type, name = "n_rows") |>
  left_join(
    df |> group_by(session_type) |> summarise(n_players = n_distinct(sensor)),
    by = "session_type"
  ) |>
  kable(caption = "Rows and unique players after cleaning") |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

---

# Data processing

## Derived KPI

We compute the **High Intensity Distance (HID)** — distance run above
15 km/h — which is the main indicator of physical exertion in football.

```{r kpi}
df <- df |>
  mutate(
    HID_km = rowSums(
      across(c(d_15_21, d_21_24, d_24_30, d_over30)),
      na.rm = TRUE
    ),
    HID_pct = if_else(distance_km > 0, HID_km / distance_km * 100, NA_real_),
    sprint_km = d_over30
  )
```

## Descriptive statistics

```{r descriptives}
desc <- df |>
  group_by(session_type) |>
  summarise(
    n           = n(),
    dist_mean   = mean(distance_km, na.rm = TRUE),
    dist_sd     = sd(distance_km, na.rm = TRUE),
    dist_median = median(distance_km, na.rm = TRUE),
    HID_mean    = mean(HID_km, na.rm = TRUE),
    HID_sd      = sd(HID_km, na.rm = TRUE),
    vmax_mean   = mean(max_speed, na.rm = TRUE),
    vmax_sd     = sd(max_speed, na.rm = TRUE),
    .groups = "drop"
  )

desc |>
  kable(digits = 2, caption = "Descriptive statistics by session type") |>
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

---

# Data visualisation

## Distance distribution — game vs practice

```{r boxplot-distance, fig.height=5}
ggplot(df, aes(x = session_type, y = distance_km, fill = session_type)) +
  geom_violin(alpha = 0.4, trim = TRUE) +
  geom_boxplot(width = 0.2, outlier.size = 1.2, outlier.alpha = 0.5) +
  scale_fill_manual(values = c("game" = "#E74C3C", "practice" = "#3498DB")) +
  labs(
    title   = "Distance distribution: match vs training",
    x       = NULL,
    y       = "Distance (km)",
    fill    = "Session type"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

## Distribution by position

```{r boxplot-position, fig.height=6}
# Keep positions with at least 20 observations
valid_pos <- df |>
  count(position) |>
  filter(n >= 20) |>
  pull(position)

df |>
  filter(position %in% valid_pos) |>
  ggplot(aes(x = reorder(position, distance_km, median),
             y = distance_km,
             fill = session_type)) +
  geom_boxplot(outlier.size = 0.8, outlier.alpha = 0.4, position = position_dodge(0.75)) +
  scale_fill_manual(values = c("game" = "#E74C3C", "practice" = "#3498DB")) +
  labs(
    title  = "Distance by position and session type",
    x      = "Position",
    y      = "Distance (km)",
    fill   = "Session type"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

## Speed zone profiles

```{r speed-zones, fig.height=5}
zones_long <- df |>
  filter(!is.na(session_type)) |>
  group_by(session_type) |>
  summarise(across(c(d_0_5, d_5_10, d_10_15, d_15_21, d_21_24, d_24_30, d_over30),
                   \(x) mean(x, na.rm = TRUE))) |>
  pivot_longer(-session_type, names_to = "zone", values_to = "mean_km") |>
  mutate(zone = factor(zone,
    levels = c("d_0_5","d_5_10","d_10_15","d_15_21","d_21_24","d_24_30","d_over30"),
    labels = c("0–5","5–10","10–15","15–21","21–24","24–30",">30")))

ggplot(zones_long, aes(x = zone, y = mean_km, fill = session_type)) +
  geom_col(position = "dodge", colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = c("game" = "#E74C3C", "practice" = "#3498DB")) +
  labs(
    title  = "Mean distance per speed zone",
    x      = "Speed zone (km/h)",
    y      = "Mean distance (km)",
    fill   = "Session type"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

## Load evolution over time (team average)

```{r time-series, fig.height=5}
ts_data <- df |>
  group_by(session_date, session_type) |>
  summarise(
    mean_dist = mean(distance_km, na.rm = TRUE),
    mean_HID  = mean(HID_km,      na.rm = TRUE),
    .groups   = "drop"
  )

ggplot(ts_data, aes(x = session_date, y = mean_dist, colour = session_type)) +
  geom_line(linewidth = 0.8, alpha = 0.7) +
  geom_point(size = 1.5, alpha = 0.8) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 1.2, alpha = 0.15) +
  scale_colour_manual(values = c("game" = "#E74C3C", "practice" = "#3498DB")) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "2 months") +
  labs(
    title   = "Team average distance over the season",
    x       = NULL,
    y       = "Mean distance (km)",
    colour  = "Session type",
    caption = "Smoothing: LOESS — shaded area = 95 % CI"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

## QQ-plots — normality check

Before choosing between parametric and non-parametric tests we
check the normality of the distance variable.

```{r qqplot, fig.height=5}
par(mfrow = c(1, 2))

dist_game     <- na.omit(df$distance_km[df$session_type == "game"])
dist_practice <- na.omit(df$distance_km[df$session_type == "practice"])

qqnorm(dist_game,     main = "QQ-plot — Game",     col = "#E74C3C", pch = 1)
qqline(dist_game,     col  = "#C0392B", lwd = 2)

qqnorm(dist_practice, main = "QQ-plot — Practice", col = "#3498DB", pch = 1)
qqline(dist_practice, col  = "#2980B9", lwd = 2)

par(mfrow = c(1, 1))
```

The QQ-plots show departures from normality (heavy tails),
which justifies the use of non-parametric tests below.

---

# Statistical analysis

## Shapiro–Wilk normality test

```{r shapiro}
# Limit to 5000 obs (Shapiro requirement)
set.seed(42)
sw_game <- shapiro.test(sample(dist_game, min(length(dist_game), 5000)))
sw_prac <- shapiro.test(sample(dist_practice, min(length(dist_practice), 5000)))

tibble(
  group   = c("game", "practice"),
  W       = c(sw_game$statistic, sw_prac$statistic),
  p_value = c(sw_game$p.value,   sw_prac$p.value)
) |>
  mutate(normal = if_else(p_value > 0.05, "yes", "no")) |>
  kable(digits   = 4,
        caption  = "Shapiro–Wilk test (H0: data are normally distributed)") |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

Both distributions are **non-normal** (p < 0.05). We therefore use
the **Wilcoxon rank-sum test** (equivalent to Mann–Whitney U).

## Wilcoxon test: game vs practice

```{r wilcoxon-distance}
wt <- wilcox.test(distance_km ~ session_type, data = df, conf.int = TRUE)

# Effect size (r = Z / sqrt(N))
wt_es <- df |>
  wilcox_effsize(distance_km ~ session_type)

tibble(
  test      = "Wilcoxon rank-sum",
  variable  = "distance_km",
  W         = wt$statistic,
  p_value   = wt$p.value,
  CI_low    = wt$conf.int[1],
  CI_high   = wt$conf.int[2],
  r_effect  = wt_es$effsize,
  magnitude = wt_es$magnitude
) |>
  kable(digits  = 4,
        caption = "Wilcoxon test: distance game vs practice") |>
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

```{r wilcoxon-multi}
# Same test on HID and max_speed
vars <- c("HID_km", "max_speed", "acc_3mss")

results <- lapply(vars, function(v) {
  d <- df |> filter(!is.na(.data[[v]]))
  wt <- wilcox.test(as.formula(paste(v, "~ session_type")), data = d)
  es <- d |> wilcox_effsize(as.formula(paste(v, "~ session_type")))
  tibble(variable  = v,
         W         = wt$statistic,
         p_value   = wt$p.value,
         r_effect  = es$effsize,
         magnitude = es$magnitude)
}) |> bind_rows()

results |>
  mutate(p_adj = p.adjust(p_value, method = "BH")) |>
  kable(digits = 4,
        caption = "Wilcoxon tests — multiple KPIs (Benjamini–Hochberg correction)") |>
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

## Kruskal–Wallis test: distance across positions

```{r kruskal}
df_pos <- df |> filter(position %in% valid_pos)

kt <- kruskal.test(distance_km ~ position, data = df_pos)

cat(sprintf("Kruskal-Wallis: H(%.0f) = %.2f, p = %.4f\n",
            kt$parameter, kt$statistic, kt$p.value))
```

```{r dunn-posthoc}
# Post-hoc Dunn test (Bonferroni correction)
dunn <- df_pos |>
  dunn_test(distance_km ~ position, p.adjust.method = "bonferroni") |>
  filter(p.adj < 0.05) |>
  arrange(p.adj)

dunn |>
  select(group1, group2, statistic, p, p.adj, p.adj.signif) |>
  kable(digits = 4,
        caption = "Post-hoc Dunn test — significant pairs only (α = 0.05)") |>
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

---

# Linear regression model

## Model specification

We model **distance** as a function of **max speed**, **HID**, and
**session type**. This tests whether the intensity variables (speed,
high-intensity distance) are predictive of total workload.

```{r lm}
df_model <- df |>
  filter(!is.na(distance_km), !is.na(max_speed), !is.na(HID_km)) |>
  mutate(session_type = factor(session_type, levels = c("practice", "game")))

model <- lm(distance_km ~ max_speed + HID_km + session_type, data = df_model)

summary(model)
```

```{r lm-tidy}
# Clean coefficient table
broom::tidy(model, conf.int = TRUE) |>
  kable(digits = 3,
        caption = "Linear model coefficients") |>
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

## Residuals diagnostics

```{r diagnostics, fig.height=8}
par(mfrow = c(2, 2))
plot(model, which = 1:4, col = alpha("#2C3E50", 0.4), pch = 1)
par(mfrow = c(1, 1))
```

```{r residuals-hist, fig.height=4}
tibble(residuals = residuals(model)) |>
  ggplot(aes(x = residuals)) +
  geom_histogram(bins = 50, fill = "#3498DB", colour = "white", alpha = 0.8) +
  geom_vline(xintercept = 0, colour = "#E74C3C", linetype = "dashed", linewidth = 1) +
  labs(title = "Distribution of residuals", x = "Residual", y = "Count") +
  theme_minimal(base_size = 13)
```

## Interpretation

```{r model-stats}
r2   <- summary(model)$r.squared
r2a  <- summary(model)$adj.r.squared
rmse <- sqrt(mean(residuals(model)^2))

cat(sprintf("R²         = %.3f\n", r2))
cat(sprintf("Adjusted R² = %.3f\n", r2a))
cat(sprintf("RMSE        = %.3f km\n", rmse))
```

---

# Outlier detection

We flag sessions where the distance deviates more than **2 standard
deviations** from the mean **for that position × session type** group.
This is a simple and operational rule for a coaching staff.

```{r outliers}
df_flagged <- df |>
  group_by(session_type, position) |>
  mutate(
    group_mean = mean(distance_km, na.rm = TRUE),
    group_sd   = sd(distance_km,   na.rm = TRUE),
    z_score    = (distance_km - group_mean) / group_sd,
    flag       = abs(z_score) > 2
  ) |>
  ungroup()

outliers <- df_flagged |>
  filter(flag) |>
  arrange(desc(abs(z_score))) |>
  select(session_date, session_type, position, sensor,
         distance_km, z_score) |>
  head(20)

outliers |>
  kable(digits = 2,
        caption = "Top 20 flagged sessions (|z| > 2 within position × type)") |>
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

```{r outlier-plot, fig.height=5}
df_flagged |>
  filter(position %in% valid_pos) |>
  ggplot(aes(x = session_date, y = distance_km, colour = flag)) +
  geom_point(aes(size = flag), alpha = 0.6) +
  scale_colour_manual(values = c("FALSE" = "#95A5A6", "TRUE" = "#E74C3C"),
                      labels = c("Normal", "Outlier (|z| > 2)")) +
  scale_size_manual(values = c("FALSE" = 0.8, "TRUE" = 2.5), guide = "none") +
  facet_wrap(~ session_type, ncol = 1) +
  labs(
    title  = "Flagged sessions over time",
    x      = NULL,
    y      = "Distance (km)",
    colour = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

---

# Summary and discussion

```{r summary-table}
df |>
  group_by(session_type, position) |>
  summarise(
    n            = n(),
    dist_median  = median(distance_km, na.rm = TRUE),
    HID_median   = median(HID_km,      na.rm = TRUE),
    vmax_median  = median(max_speed,   na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(n >= 10) |>
  arrange(session_type, desc(dist_median)) |>
  kable(digits = 2,
        caption = "Dashboard: median KPIs by session type and position (n ≥ 10)") |>
  kable_styling(bootstrap_options = c("striped", "condensed", "hover")) |>
  pack_rows("Game", 1, sum(df |> filter(session_type=="game",
             position %in% (df |> count(session_type, position) |>
             filter(n >= 10) |> pull(position))) |>
             distinct(position) |> nrow())) |>
  row_spec(0, bold = TRUE)
```

## Main findings

**1. Game vs practice load (Wilcoxon test)**  
The total distance in matches is significantly higher than in training
(large effect size). The same pattern holds for HID and max speed,
confirming that GPS data correctly captures the difference in
intensity between competition and training contexts.

**2. Position-specific differences (Kruskal–Wallis + Dunn)**  
Load differs significantly across positions. Wide players and
midfielders cover the greatest distances, while goalkeepers show
distinctly lower values — as expected biomechanically.

**3. Linear model (R² ≈ `r round(summary(model)$r.squared, 2)`)**  
Max speed and HID explain a large portion of the variance in total
distance. Session type adds independent predictive power.
The residuals diagnostics show mild heteroscedasticity at high
predicted values, suggesting the model could be improved with a
log-transformation or by stratifying by position.

**4. Outlier detection**  
The z-score approach identified sessions with anomalous loads. Before
using these alerts operationally, each flagged session should be
cross-checked against context (injury return, overload week, sensor
error).

## Limitations and perspectives

- The quality rules (thresholds) were calibrated on reasonable domain
  knowledge; a more rigorous approach would use cross-validation or
  external validation data.  
- The linear model assumes independence of observations; mixed-effects
  models accounting for repeated measures per player would be more
  appropriate.  
- Tracking raw data (position/speed time series) could enable
  intra-session analyses (e.g., fatigue in the second half) not
  accessible from the aggregated summary.

---
